Tutorial_2

### 2024-09-12 Tutorial_2_Bayesian/Tutorial_2.Rmd
update tutorial 3

In this file we build Bayesian hazard regression models for the first experiment of Panis and Schmidt (2016) using only the no-mask trials (prime = blank, congruent, or incongruent), and visualize the posterior distributions of the effects of congruent and incongruent prime types, relative to blank prime trials.

@@

```{r data-M1}
# remove unnecessary columns before fitting a model
M1_data <- ptb_data %>% select(-c(bl,tr,trial,period, period_9,d9)) # 12840 obs. of 14 variables
head(M1_data)
summary(M1_data)

#M1_data <- within(M1_data, condition <- relevel(condition, ref = "blank"))

@@

```{r check-model-M1}

summary(model_M1)

model_M1c <- readRDS("Tutorial_2_Bayesian/models/model_M1c.rds")

fixef(model_M1)

```{r data-M1b}
# remove unnecessary columns before fitting a model
M1b_data <- ptb_data %>% select(-c(bl,tr,trial,period, period_9,d9)) # 12840 obs. of 14 variables
head(M1b_data)
summary(M1b_data)

#M1_data <- within(M1_data, condition <- relevel(condition, ref = "blank"))

```{r priors-cloglog-M1}
priors_M1b <- c(
  set_prior("skew_normal(-0.2,0.71,-2.2)", class = "b"),       # weakly informative prior for beta parameters when using cloglog link
  #set_prior("skew_normal(-1,1,-2)", class = "Intercept"), # weakly informative prior for alpha parameters when using cloglog link
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "d6"), 
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "d7"), 
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "d8"), 
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "d10"), 
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "d11"), 
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "d12"), 
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "d13"), 
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "d14"), 
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "d15"),
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),
  set_prior("normal(0, 1)", class = "sd"),                     # prior for standard deviation of random effects
  set_prior("lkj(2)", class = "cor")                           # prior for correlations between random effects
)
```

model_M1b <-
   brm(data = M1b_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + d6 + d7 + d8 + d10 + d11 + d12 + d13 + d14 + d15 + 
         condition + trial_c +

                   (0 + Intercept + d6 + d7 + d8 + d10 + d11 + d12 + d13 + d14 + d15 + condition + trial_c | pid),
       prior = priors_M1b,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_M1b")
```

117 minutes

```{r}
summary(model_M1)

fixef(model_M1b)

```{r fit-model-M1c}
plan(multicore)

model_M1c <-
   brm(data = M1b_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + d6 + d7 + d8 + d10 + d11 + d12 + d13 + d14 + d15 + 
         condition + trial_c +

                   (1 + d6 + d7 + d8 + d10 + d11 + d12 + d13 + d14 + d15 + condition + trial_c | pid),
       prior = priors_M1b,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_M1c")
```

@@

```{r check-model-M2}
summary(model_M2)
model_M2b <- readRDS("Tutorial_2_Bayesian/models/model_M2b.rds")
fixef(model_M2b)
```

```{r priors-cloglog-M2b}
priors_M2b <- c(
  set_prior("skew_normal(-0.2,0.71,-2.2)", class = "b"),       # weakly informative prior for beta parameters when using cloglog link
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),      # weakly informative prior for Intercept when using cloglog link
  set_prior("normal(0, 1)", class = "sd"),                     # prior for standard deviation of random effects
  set_prior("lkj(2)", class = "cor")                           # prior for correlations between random effects
)
```

model_M2b <-
   brm(data = M2_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + period_9 + I(period_9^2) + I(period_9^3) + 
                          condition + trial_c +
                          (1 + period_9 + I(period_9^2) + I(period_9^3) + 
                          condition + trial_c | pid),
       prior = priors_M2,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_M2b")
```

```{r}
fixef(model_M2b)
```

@@

model_M4 <- 
   brm(data = M4_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + # Note that duplicate terms in the model formula are ignored
                           condition*period_9 + 
                           condition*I(period_9^2) + 
                           condition*I(period_9^3) +
                           trial_c*period_9 + 
                           trial_c*I(period_9^2) + 
                           trial_c*I(period_9^3) +
                           I(trial_c^2)*period_9 + 
                           I(trial_c^2)*I(period_9^2) + 
                           condition*period_9*trial_c + 
                           condition*I(period_9^2)*trial_c +
                           condition*period_9*I(trial_c^2) + 
                           condition*I(period_9^2)*I(trial_c^2) +
                           (1 + condition*period_9 +
                           condition*I(period_9^2) +
                           condition*I(period_9^3) +
                           trial_c*period_9 + 
                           trial_c*I(period_9^2) + 
                           trial_c*I(period_9^3) +
                           I(trial_c^2)*period_9 + 
                           I(trial_c^2)*I(period_9^2) + 
                           condition*period_9*trial_c + 
                           condition*I(period_9^2)*trial_c +
                           condition*period_9*I(trial_c^2) + 
                           condition*I(period_9^2)*I(trial_c^2)  | pid),
       prior = priors_M4,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_M4")
```

@@

@@

Plot Pareto k estimates

@@

# visualize posterior distributions of the effects of congruent and incongruent primes on cloglog-hazard relative to blank prime, for each time bin in trial 1000.

@@

```{r}
post <-as_draws_df(model_M4) %>% # 8000 draws x 715 variables
   select(starts_with("b_")) %>%       # 8000 x 31
   expand_grid(period_9 = -3:6) %>%   
   mutate(period_9sq = period_9^2,
          period_9cu = period_9^3,
          trial100 = -900/1000)  
# effects for trial 1000
plot_facet <- post %>% 
  mutate(Congruent = b_conditioncongruent + period_9 * `b_conditioncongruent:period_9` + period_9sq * `b_conditioncongruent:Iperiod_9E2` + period_9cu * `b_conditioncongruent:Iperiod_9E3`,
         Incongruent = b_conditionincongruent + period_9 * `b_conditionincongruent:period_9` + period_9sq * `b_conditionincongruent:Iperiod_9E2` + period_9cu * `b_conditionincongruent:Iperiod_9E3`) %>%

@@

# Plot model-based predicted cloglog-hazard for subject 6

```{r}
make_fitted <- function(fit, scale, ...) {

@@

# model_M4: Plot predicted cloglog(hazard) for single subject using fitted() in make_fitted().

@@

```{r}
selected_subject = 6 # select pid here (1:6) manually

get_variables(model_M4)
```

First, on the cloglog-scale:

```{r}
make_fitted(model_M4, scale = "linear") %>% 
  mutate(cond   = rep(rep(c("N","C","I"),each=30),6),
         trial  = factor(trial_c,levels=c(-0.5,0.0,0.5),labels=c("trial 500","trial 1000","trial 1500")),
         period = period_9+9) %>% 
  filter(pid==selected_subject) %>% 
  # plot!
  ggplot(aes(x = period, y = Estimate, ymin = Q2.5, ymax = Q97.5,
             fill = condition, color = condition)) +
  geom_ribbon(alpha = 1/5, linewidth = 0) +
  geom_line() +
  scale_fill_viridis_d(NULL, option = "A", end = .6, direction = -1) +
  scale_color_viridis_d(NULL, option = "A", end = .6, direction = -1) +
  scale_x_continuous("time bin endpoint", breaks = 6:15, limits = c(6, 15), labels=c(6:15)*40) +
  ylab("fitted cloglog(hazard)") +
  coord_cartesian(ylim = c(-5, 1)) +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill="white",color="black"),
        legend.position = "top",
        axis.text.x = element_text(angle=90)) +
  facet_wrap(~ trial)
```

Same on hazard scale

```{r}
make_fitted(model_M4, scale = "response") %>% 
  mutate(cond   = rep(rep(c("N","C","I"),each=30),6),
         trial  = factor(trial_c,levels=c(-0.5,0.0,0.5),labels=c("trial 500","trial 1000","trial 1500")),
         period = period_9+9) %>% 
  filter(pid==selected_subject) %>% 
  # plot!
  ggplot(aes(x = period, y = Estimate, ymin = Q2.5, ymax = Q97.5,
             fill = condition, color = condition)) +
  geom_ribbon(alpha = 1/5, linewidth = 0) +
  geom_line() +
  scale_fill_viridis_d(NULL, option = "A", end = .6, direction = -1) +
  scale_color_viridis_d(NULL, option = "A", end = .6, direction = -1) +
  scale_x_continuous("time bin endpoint", breaks = 6:15, limits = c(6, 15), labels=c(6:15)*40) +
  ylab("fitted hazard") +
  coord_cartesian(ylim = c(0, 1)) +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill="white",color="black"),
        legend.position = "top",
        axis.text.x = element_text(angle=90)) +
  facet_wrap(~ trial)
```

##########

Fit a model that relaxes proportionality assumption.

```{r}
# remove unnecessary columns before fitting a model
ptb_data <- ptb_data %>% select(-c(trial_c)) # 12840 obs. of 4 variables
head(ptb_data,n=17)
summary(ptb_data)
```

```{r cubic-random-effects-model}
plan(multicore)

model_cubic_RE <-
   brm(data = ptb_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + 
                           condition*period_9 + 
                           condition*I(period_9^2) + 
                           condition*I(period_9^3) +
                           (1 + condition*period_9 +
                           condition*I(period_9^2) +
                           condition*I(period_9^3) | pid),
       prior = priors,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_cubic_RE")

# the formula above comes down to the following effects:
# cloglog[h(t)]=Intercept+ ##the estimated cloglog[h(t)] in bin 9 for trial 1000 for reference condition "blank"
# period_9+I(period_9^2)+I(period_9^3)+ ##a cubic specification of TIME for reference condition "blank"
# congruent+congruent:period_9+congruent:I(period_9^2)+congruent:I(period_9^3)+ ##effect of congruent interacts cubicly with TIME
# incongruent+incongruent:period_9+incongruent:I(period_9^2)+incongruent:I(period_9^3) ##effect of incongruent interacts cubicly with TIME
```

```{r check-model-cubic-RE}
summary(model_cubic_RE)
```

#### relax all ass

```{r cubictime-quadtrial-model}
plan(multicore)

model_cubictime_quadtrial <-
   brm(data = ptb_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + 
                           condition*period_9*trial_c + 
                           condition*period_9*I(trial_c^2) + 
                           condition*I(period_9^2)*trial_c + 
                           condition*I(period_9^2)*I(trial_c^2) +
                           condition*I(period_9^3) +
                      (1 + condition*period_9*trial_c +
                           condition*period_9*I(trial_c^2) +
                           condition*I(period_9^2)*trial_c +
                           condition*I(period_9^2)*I(trial_c^2) +
                           condition*I(period_9^3) | pid),
       prior = priors,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_cubictime_quadtrial")

# the formula above comes down to the following effects:
# cloglog[h(t)]=Intercept+ ##the estimated cloglog[h(t)] in bin 9 for trial 1000 for reference condition "blank"
# period_9+I(period_9^2)+I(period_9^3)+ ##a cubic specification of TIME for reference condition "blank"
# congruent+congruent:period_9+congruent:I(period_9^2)+congruent:I(period_9^3)+ ##effect of congruent interacts cubicly with TIME
# incongruent+incongruent:period_9+incongruent:I(period_9^2)+incongruent:I(period_9^3)+ ##effect of incongruent interacts cubicly with TIME
# trial_c + trial_c:period_9 + ##linear effect of trial number in bin 9 can change linearly over TIME for "blank"
# trial_c:congruent + trial_c:incongruent + ##linear effect of trial number in bin 9 can differ for congruent and incongruent
# trial_c:congruent:period_9 + trial_c:incongruent:period_9 ##how the linear effect of trial number changes linearly over TIME can differ for congruent and incongruent
```

# plot hazard functions

```{r}
model_full_RE <- readRDS("Tutorial_2_Bayesian/models/model_full_RE.rds") 
model_full_RE_newprior <- readRDS("Tutorial_2_Bayesian/models/model_full_RE_newprior.rds") 
fixef(model_full_RE)
fixef(model_full_RE_newprior)
```

# visualize posterior distributions of the effects of congruent and incongruent primes on hazard relative to blank prime, for each time bin in trial 1000, and for each time bin in trial 100.

```{r load-model}
model_full_RE <- readRDS("Tutorial_2_Bayesian/models/model_full_RE.rds") 
```

```{r make-plot-effects}
post <-as_draws_df(model_full_RE) %>% # 8000 draws x 299 variables
   select(starts_with("b_")) %>%       # 8000 x 18
   expand_grid(period_9 = -3:6) %>%   
   mutate(period_9sq = period_9^2,
          period_9cu = period_9^3,
          trial100 = -900/1000)  
# effects for trial 1000
p1congruent <- post %>% 
  mutate(postdistr = b_conditioncongruent + period_9 * `b_conditioncongruent:period_9` + period_9sq * `b_conditioncongruent:Iperiod_9E2` + period_9cu * `b_conditioncongruent:Iperiod_9E3`) %>%
  group_by(period_9) %>%

  ggplot(aes(x = period_9, y = postdistr)) +
  stat_lineribbon(show.legend=F) +
  scale_fill_brewer() +
  geom_hline(yintercept=0, linetype="dashed", color = "red") +
  scale_x_continuous(breaks = c(-3:6), labels=c(((-3:6)+9)*40),
                     limits = c(-3,6)) +
  scale_y_continuous(breaks = c(-6,-4,-2,0,2,4,6), limits = c(-6,6)) +
  labs(title = "trial 1000", x = "time bin", y = "effect of congruent") +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90))

p1congruent + p2incongruent
#(p1+p2+p3)/(p4+p5+p6)/(p7+p8+p9)/(p10+p11+p12)/(p13+p14+p15)
```

```{r save-plot-effects}
ggsave("figures/effects_knot_1.png", width = 12, height = 16, dpi = 600)

>>>>

In this file we plot priors for the logit and cloglog link functions, build Bayesian hazard regression models for the first experiment of Panis and Schmidt (2016) using only the no-mask trials (prime = blank, congruent, or incongruent), and visualize the posterior distributions of the effects of congruent and incongruent prime types, relative to blank prime trials for the selected model.

@@

```{r data-M1}
# remove unnecessary columns before fitting a model
M1_data <- ptb_data %>% select(-c(bl,tr,trial,period, period_9,d9)) # 12840 obs. 
head(M1_data)
summary(M1_data)
```

@@

```{r check-model-M1}
summary(model_M1)
fixef(model_M1)
```

@@

```{r check-model-M2}
summary(model_M2)
fixef(model_M2)
```

@@

model_M4 <- 
   brm(data = M4_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + # Note that duplicate terms in the model formula are ignored
                          condition*period_9*trial_c + 
                          condition*period_9*I(trial_c^2) + 
                          condition*I(period_9^2)*trial_c +
                          condition*I(period_9^2)*I(trial_c^2) +
                          condition*I(period_9^3) +
                          trial_c*I(period_9^3) +
                          (1 +  condition*period_9*trial_c +
                                condition*period_9*I(trial_c^2) + 
                                condition*I(period_9^2)*trial_c +
                                condition*I(period_9^2)*I(trial_c^2)  
                                condition*I(period_9^3) +
                                trial_c*I(period_9^3) | pid),
       prior = priors_M4,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_M4")
```

@@

# the formula above comes down to the following effects:
# cloglog[h(t)]=Intercept+     ## the estimated cloglog[h(t)] in bin 9 for trial 1000 for reference condition "blank"
# period_9+I(period_9^2)+I(period_9^3)+    ##a cubic specification of TIME for reference condition "blank"
# congruent+congruent:period_9+congruent:I(period_9^2)+congruent:I(period_9^3)+      ##effect of congruent interacts cubicly with TIME
# incongruent+incongruent:period_9+incongruent:I(period_9^2)+incongruent:I(period_9^3)+     ##effect of incongruent interacts cubicly with TIME
# trial_c + trial_c:period_9 + ##linear effect of trial number in bin 9 can change linearly over TIME for "blank"
# trial_c:congruent + trial_c:incongruent + ##linear effect of trial number in bin 9 can differ for congruent and incongruent
# trial_c:congruent:period_9 + trial_c:incongruent:period_9 ##how the linear effect of trial number changes linearly over TIME can differ for congruent and incongruent
```

@@

## Pareto k estimates

@@

## Visualize posterior distributions of the effects of congruent and incongruent primes on cloglog-hazard relative to blank prime, for each time bin in trial 1000.

@@

```{r}
post <-as_draws_df(model_M4) %>% # 8000 draws x 715 variables
   select(starts_with("b_")) %>%       # 8000 x 31
   expand_grid(period_9 = -3:6) %>%   
   mutate(period_9sq = period_9^2,
          period_9cu = period_9^3,
          trial100 = -900/1000)

# effects for trial 1000
plot_facet <- post %>% 
  mutate(Congruent = b_conditioncongruent + period_9 * `b_conditioncongruent:period_9` + period_9sq * `b_conditioncongruent:Iperiod_9E2` + period_9cu * `b_conditioncongruent:Iperiod_9E3`,
         Incongruent = b_conditionincongruent + period_9 * `b_conditionincongruent:period_9` + period_9sq * `b_conditionincongruent:Iperiod_9E2` + period_9cu * `b_conditionincongruent:Iperiod_9E3`) %>%

@@

# Plot model-based predicted cloglog-hazard and hazard functions, for a single subject

```{r select-pid}
selected_subject = 6 # select pid here (1:6) manually
```

```{r help-function}
make_fitted <- function(fit, scale, ...) {

@@

# Set up new data

@@

# First, on the cloglog-scale:

```{r}
make_fitted(model_M4, scale = "linear") %>% 
  mutate(cond   = rep(rep(c("N","C","I"),each=30),6),
         trial  = factor(trial_c,levels=c(-0.5,0.0,0.5),labels=c("trial 500","trial 1000","trial 1500")),
         period = period_9+9) %>% 
  filter(pid==selected_subject) %>% 
  # plot
  ggplot(aes(x = period, y = Estimate, ymin = Q2.5, ymax = Q97.5,
             fill = condition, color = condition)) +
  geom_ribbon(alpha = 1/5, linewidth = 0) +
  geom_line() +
  scale_fill_viridis_d(NULL, option = "A", end = .6, direction = -1) +
  scale_color_viridis_d(NULL, option = "A", end = .6, direction = -1) +
  scale_x_continuous("time bin endpoint", breaks = 6:15, limits = c(6, 15), labels=c(6:15)*40) +
  ylab("fitted cloglog(hazard)") +
  coord_cartesian(ylim = c(-5, 1)) +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill="white",color="black"),
        legend.position = "top",
        axis.text.x = element_text(angle=90)) +
  facet_wrap(~ trial)
```

# Second, on hazard scale:

```{r}
make_fitted(model_M4, scale = "response") %>% 
  mutate(cond   = rep(rep(c("N","C","I"),each=30),6),
         trial  = factor(trial_c,levels=c(-0.5,0.0,0.5),labels=c("trial 500","trial 1000","trial 1500")),
         period = period_9+9) %>% 
  filter(pid==selected_subject) %>% 
  # plot
  ggplot(aes(x = period, y = Estimate, ymin = Q2.5, ymax = Q97.5,
             fill = condition, color = condition)) +
  geom_ribbon(alpha = 1/5, linewidth = 0) +
  geom_line() +
  scale_fill_viridis_d(NULL, option = "A", end = .6, direction = -1) +
  scale_color_viridis_d(NULL, option = "A", end = .6, direction = -1) +
  scale_x_continuous("time bin endpoint", breaks = 6:15, limits = c(6, 15), labels=c(6:15)*40) +
  ylab("fitted hazard") +
  coord_cartesian(ylim = c(0, 1)) +
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill="white",color="black"),
        legend.position = "top",
        axis.text.x = element_text(angle=90)) +
  facet_wrap(~ trial)
```
